Gradual Soundness: Lessons from Static Python

### 2022-05-19 pj-7.1/programming-camera/paper.tex
pj: prepare src bundle for tobias

\begin{itemize}
  \item
    \emph{Evidence}.
    We present evidence that \SP{} improves performance with few code changes.
    At Instagram, the web server has become significantly more efficient
    after the application of \SP{} to high-profile modules.
    Because the server code is closed-source, we additionally present data for
    \numbenchmark{} public microbenchmarks~(\cref{s:microbenchmarks}).
    Using only shallow types, \SP{} runs slightly faster than Python
    despite the costs of enforcing soundness~(avg. \pctnum{20}).
    With fine-tuned types, \SP{} consistently outperforms the Python
    baseline~(avg. \pctnum{70}).

@@

Surface types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We choose to write the dynamic type as $\sptdyn$ for simplicity; technically,
\SP{} does not have a surface syntax for its dynamic
type and instead allows missing and unknown types~(\cref{s:type-dynamic}).
We assume that all unions are written in a flat and simplified form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be normalized to $\sptdyn$.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython had improved performance by compiling these modules to C,
but its partial adoption also led to an awkward workflow with a
few compiled modules spread across an interpreted codebase.
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% improvement in web server CPU efficiency.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

@@

Because the \SP{} team has been changing code as well as types during its
migrations, the question arises as to whether the changes themselves
are responsible for more speedups than the types and gradual soundness.
%% Carl: "I reviewed the ~30 conversion diffs"
We therefore reviewed conversion diffs that significantly improved
performance (\nnum{30}) to assess the extent of the code changes.
There are nine common kinds of changes: five that satisfy the type checker
and four that aim to improve performance.

@@

With its mix of shallow and concrete types, \SP{} is one of a few gradual
language that lets programmers choose between multiple type-enforcement strategies.
Thorn and StrongScript offer a choice of concrete types and optional (unsound)
\emph{like} types.
Typed Racket supports deep (structural) types, shallow types, and optional
types~\cite{g-pldi-2022}.
By contrast to these others, \SP{} asks programmers to change their code
in way that reflects changes to the types~(\cref{s:checked-type}).
The changes impose a maintenance burden, but enable performance improvements and low-cost checks.

@@

Benchmarks are run on a 64-bit CentOS Linux system, using a build of Cinder
compiled with default configure settings. Overall process user time is measured
via the \code{time} utility. Each benchmark/configuration is run 10 times (after an
initial warmup run to update the bytecode cache file and ignoring compilation time),
and all individual data points are
recorded. All data points are in seconds.

>>>>

\begin{itemize}
  \item
    \emph{Evidence}.
    We present evidence that \SP{} improves performance with few code changes
    / refactorings.
    The Instagram web server has become significantly more efficient
    thanks to a gradual application of \SP{} to high-profile modules.
    Because the server code is closed-source, we additionally present data for
    four public microbenchmarks~(\cref{s:microbenchmarks}).
    Using only shallow types, \SP{} runs slightly faster than Python
    despite the costs of enforcing soundness~(avg. \pctnum{20}).
    With fine-tuned types, \SP{} consistently outperforms the Python
    baseline~(avg. \pctnum{70}).

@@

Surface types $\sptype$ include the dynamic type, types for the basic values, one type $\sptclass$ for every
user-defined class, and union types.
We choose to write the dynamic type as $\sptdyn$ for simplicity; technically,
\SP{} does not have a surface syntax for its dynamic
type and instead allows missing types~(\cref{s:type-dynamic}).
We assume that all unions are written in a flat and simplified form, e.g., that
$\sptunion{\sptint, \sptunion{\sptdyn, \sptclass_0}}$ would be normalized to $\sptdyn$.

@@

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython had improved performance by compiling these modules to C,
but its partial adoption led to an awkward workflow with a
few compiled modules spread across an interpreted codebase.
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% improvement in web server CPU efficiency.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

@@

Because the \SP{} team has been changing code as well as types during its
migrations, the question arises as to whether the refactoring caused
more speedups than the types and gradual soundness.
%% Carl: "I reviewed the ~30 conversion diffs"
We therefore reviewed 30 conversion diffs that significantly improved
performance to assess the extent of the code changes.
Across the diffs, there are nine common kinds of changes: five that satisfy the
type checker and four that aim to improve performance.

@@

With its mix of shallow and concrete types, \SP{} is one of a few gradual
language that lets programmers choose between multiple type-enforcement strategies.
Thorn and StrongScript offer a choice of concrete types and optional (unsound)
\emph{like} types.
Typed Racket supports deep (structural) types, shallow types, and optional
types~\cite{g-pldi-2022}.
By contrast to these others, \SP{} asks programmers to change their code
in way that reflects changes to the types~(\cref{s:checked-type}).
The changes impose a maintenance burden, but enable performance improvements
and an efficient implementation of run-time checks.

@@

Benchmarks are run on a 64-bit CentOS Linux system, using a build of Cinder
compiled with default configure settings. Overall process user time is measured
via the \code{time} utility. Each benchmark/configuration is run 10 times (after an
initial warmup run to update the bytecode cache file and thereby ignore compilation time),
and all individual data points are
recorded. All data points are in seconds.
